Dissertation Proposal

### 2023-10-26 paper.tex
line numbers
rqs and future plan

@@

\begin{document}
\maketitle

@@

Unfortunately, this combination of features appears to also be non-trivial for
programmers to understand. Consider the following scenario: to create a
calculator, we have to construct a callback that can be attached to
each button. The following Python program seems to achieve the
construction of these callbacks and the act of pushing the buttons in
order:
\newpage
\begin{lstlisting}[language=Python]
button_list = []

@@

In response, we have (a) run a multi-year, multi-phase study to
identify these misunderstandings while trying to work around the blind
spots of experts (and also drawing on prior work),
and (b) constructed the SMoL Tutor, an interactive tutor to address
these misunderstandings. The SMoL Tutor is not a \emph{programming} tutor;
rather, it assumes the user has some basic facility with programming
(i.e., that they are already familiar with concepts like vectors,
mutation, and higher-order functions). It is built around SMoL's
features and their interactions, and draws on cognitive and
educational psychology concepts to expressly identify
and correct misconceptions.

Concretely, we make the following contributions:
\begin{enumerate}

@@

  \item We present a tutoring system that measurably corrects these
        misconceptions.

@@

@@

Nevertheless, most SMoL programs are easy to translate to other
languages. We present machine-translated Python and JavaScript
versions of our programs in \supplRef{Translated\_Programs.html}. Furthermore, we intend to
make a multi-lingual tutor in the future (\Cref{s:discussion}).

@@

@@

The other major concern is the tie to Lispy syntax. Learners in other
settings may do worse or even better with it. Building a Tutor that
supports multiple, and more traditional, syntaxes should help address
this issue. We defer this to future work.

\section{Discussion}%
\label{s:discussion}

The paper has already identified several areas for future work:
\begin{itemize}
\item testing on more varied populations;
\item using other syntaxes;
\item having more questions that uniquely identify minsconceptions;
  and,
\item enabling textual responses in the Tutor so we can better
  characterize wrong answers.
\end{itemize}
These can help get us even closer to a good approximation of a true concept
inventory. We would also like the Tutor to make more use of the education
theories discussed in \Cref{s:background}.

We focus here on some issues that we think warrant broader
discussion.

\paragraph{The State-Aliasing-Function Triangle}

Our problematic programs hardly include any ``advanced'' programming
features: there are no threads, asynchrony, sophisticated type systems, ownership,
inversion of control, etc. Indeed, many of those features typically
build on an understanding of these basics (e.g., it is hard to make
sense of ownership \cite{clarkeOwnershipTypesFlexible1998} without a good understanding of this
triangle). But many populations seem to struggle even with this, which
may explain why concepts like ownership are considered hard \cite{crichtonGroundedConceptualModel2023}.

It is worth noting that we find problems even without all three
components, as \Cref{s:smol-quizzes} shows! Nevertheless, we think it
would be useful for curricula to focus on achieving mastery of this
triangle. This may require a deep revision of widely accepted
pedagogy. For instance, it is common to explain variables as
``boxes''. But if taken seriously by a learner, this metaphor may
cause more harm than good. As prior research \cite{putnamSummaryMisconceptionsHigh1986,groverMeasuringStudentLearning2017,hermansThinkingOutBox2018} shows, students expect that a variable can then contain more
than one value, removing one makes the other accessible, etc.  That
research has not explored aliasing, but the metaphor may affect that
too. A box is a closed object, so a (mutable) value put ``into'' it
clearly can't be modified by another ``box'' (variable)---i.e., it not
only doesn't explain but is antithetical to aliasing.

\paragraph{Terminology}

It is common to teach programming using the terms ``call-by-value''
and ``call-by-reference''. The reader will note that we instead have
\emph{three} \miscon{ByRef} misconceptions.

The emphasis on ``calling'' suggests that the semantics is associated
\emph{only} with calls. That leaves open what happens when one simply
binds a variable. In a reasonable language (and definitionally, in
SMoL), the behavior is exactly the same: the formal parameter can be
viewed as a binding to the value of the actual. But students often
form inconsistent views because the ``call'' terminology breaks this
deep similarity. We therefore recommend that languages in general use
the term \emph{bind-by-}, to emphasize the underlying semantic unity
of these syntactically different mechanisms.

We feel that further confusion is caused by terms like ``pass'' and
``return''. We often vocalize the call \verb|(f x)| as ``passing
\verb|x| to \verb|f|''; saying ``passing \emph{the value of} \verb|x|
to \verb|f|'' is a mouthful. But is it then surprising that students
think \verb|x| is being aliased? Similarly, consider a statement like
\verb|return y| (in Python syntax). Of course, semanticists understand
that it is the \emph{value} of \verb|y|, not \verb|y| itself, that is
being returned. Nevertheless, it is not surprising if this also
results in assumptions that \verb|y| is either aliased or that it has
escaped from the function (leading to an interpretation of dynamic
scope). We believe there is a need for significant research to
investigate these kinds of effects, which are similar to but not
strictly the same as ``vernacular misconceptions'' \cite{nationalresearchcouncilScienceTeachingReconsidered1997}.

>>>>

\usepackage[switch]{lineno}
\usepackage{xcolor}
\renewcommand\thelinenumber{\textcolor{red}{\arabic{linenumber}}}

@@

\begin{document}
\maketitle
\linenumbers

@@

Unfortunately, this combination of features appears to also be non-trivial for
programmers to understand. Consider the following scenario: to create a
calculator, we have to construct a callback that can be attached to
each button. The following Python program seems to achieve the
construction of these callbacks and the act of pushing the buttons in
order:
\begin{lstlisting}[language=Python]
button_list = []

@@

This lead to my \textbf{research questions}:

\begin{description}
  \item[RQ1] What are the misconceptions that are about program
  behavior, apply to most modern programming languages, and present
  (even) in students with prior programming backgrounds?
  \item[RQ2] What would be a tutoring system (the SMoL Tutor) that
  draws on cognitive and educational psychology concepts and that aim
  at correcting the misconceptions?
  \item[RQ3] How effective is SMoL Tutor at correcting the
  misconceptions?
\end{description}

To answer my research questions, I have (a) run a multi-year, multi-phase study to
identify these misunderstandings while trying to work around the blind
spots of experts (and also drawing on prior work),
and (b) constructed the SMoL Tutor, an interactive tutor to address
these misunderstandings. The SMoL Tutor is not a \emph{programming} tutor;
rather, it assumes the user has some basic facility with programming
(i.e., that they are already familiar with concepts like vectors,
mutation, and higher-order functions). It is built around SMoL's
features and their interactions, and draws on cognitive and
educational psychology concepts to expressly identify
and correct misconceptions.

Concretely, I have made the following contributions:
\begin{enumerate}

@@

  \item We present a tutoring system, the SMoL Tutor, that measurably
        corrects these misconceptions.

@@

In the near future, I would like to

@@

Nevertheless, most SMoL programs are easy to translate to other
languages. We present machine-translated Python and JavaScript
versions of our programs in \supplRef{Translated\_Programs.html}. Furthermore, we intend to
make a multi-lingual tutor in the future (\Cref{s:future-work}).

@@

Because we don't randomize Tutor question order, it is hard to tell
whether the general decreasing trend in \cref{f:tutor-trend-plots}
shows misconceptions being corrected, or shows properties of the
questions (e.g., maybe later questions in the plots are easier because
their programs happen to be shorter).

@@

The other major concern is the tie to Lispy syntax. Learners in other
settings may do worse or even better with it. Building a Tutor that
supports multiple, and more traditional, syntaxes should help address
this issue. We defer this to future work (\cref{s:future-work}).

\section{Future Work \& Timeline}%
\label{s:future-work}

This semester (Fall 2023) I am already deploying a newer SMoL Tutor at
Brown. This new tutor is able to display programs in JavaScript and
Python, in addition to the Lispy syntax. The goal of this semester is
to compare the new data with the current results, i.e., a comparison
between uni-syntax tutor and multi-syntax tutor.

By the end of Fall 2023, I would like to improve the SMoL Tutor:
\begin{enumerate}
  \item Ask students to explained when they gave a wrong answer in a
  interepreting question. With student explanations, I check whether
  choosing the wrong answer indicates holding the corresponding misconception(s).
  \item Adjust the collection of interpreting questions such that
  every misconception has at least two questions that aims at
  correcting the misconception.
  \item Partially randomize the order of questions.
  \item Try to reduce the chance of missing values.
\end{enumerate}

Several faculties from other institutions have expressed interest in
using the Tutor in Spring 2024. My goal is to collect data from at
least two other institutions. I am officially taking a personal leave
in Spring 2024, so I do not plan to make more progress other than
data collection.

During Fall 2024, I plan to collect a new round of data from Brown. I
expect this datasets and the datasets from Spring 2024 would give a
more rigorously evaluation of the effectiveness of SMoL Tutor. The
improvements to be made in Fall 2023 should resolve many of the
threats to validity. In this semester I also want to write a new paper
about the SMoL Tutor.

I plan to defense in Spring 2025.
